The Design and Implementation of the Synchronous Language \CEU

### 2016-03-31 paper.tex
...

\begin{abstract}
\CEU is a reactive language inspired by Esterel that targets constrained 
embedded platforms and ensures safe concurrency by handling threats at compile 
time.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis and results in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
We discuss the design of \CEU and propose a formal semantics for its particular 
control mechanisms, such as parallel compositions, finalization, and internal 
events.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

% TODO: input/output declarations

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end

.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with intra-reaction communication, \CEU supports internal events which 
programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
\item No other trails awake to \code{b}.
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

% CONTRIBS
    - queue in the semantics
    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel which aims to offer a 
simpler semantics and more fine-grained control over the program execution.
%
\CEU employs an event-triggered notion of time that allows for a simple
reasoning about reactions in parallel and enables compile-time analysis to 
produce deterministic and concurrency-safe programs.
%
We discuss the particularities of our design in comparison to Esterel, such as 
stack-based internal events, temporal analysis for concurrent statements, safe 
integration with \C, and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel \ax and 
\CEU \bx for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 5--10, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 4--11), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 4--13):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
input  A, B;
output O;
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end

.
\end{lstlisting}
\centering\small{\ax Esterel}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
input  void A, B;
output void O;
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\centering\small{\bx \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

@@

Esterel makes no semantic distinctions between internal and external signals.
In particular, programs can emit external and internal signals simultaneously, 
with all coexisting during the entire reaction.
%
%In \CEU, a reaction starts from an external input event and programs cannot 
%emit inputs at all.
%Therefore, the occurring input event is unique during the entire reaction, 
%resulting in intrinsic queue-based handling.
%
In \CEU, subsequent external events coming from the environment define a 
timeline, which programs cannot interfere by emitting new external events.
%
To cope with deterministic intra-reaction communication, \CEU supports internal 
events which programs can \code{emit} and \code{await}.
In contrast with external events, internal events follow a stack-based 
execution policy similar to subroutine calls in typical programming languages.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel \ax and \CEU \bx.
%
In Esterel, when \code{A} occurs, \code{B} is emitted (ln. 4--5) and both 
events become active, resulting in the invocation of \code{f()} and \code{g()} 
in no particular order.
%
In \CEU, the occurrence of \code{A} makes the program behave as follows:
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (ln. 4), broadcasts \code{b}, and pauses.
\item 2nd trail awakes (ln. 8), calls \code{\_g()}, and terminates.
      \emph{(No other trails awake to \code{b}.)}
\item 1st trail (on top of the stack) resumes, calls \code{\_f()}, and 
    terminates.
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates.
\end{enumerate}
}

@@

% CONTRIBS
    - queue in the semantics
    - permeates all aspects
    - As far as we know, \CEU is the first Esterel-based language to HAVE?
      queue-based/unique in the semantics of the language and not in the
      library
    - stack
        - deterministic intra-reaction execution
    - imposed intra-reaction execution
        - for inter-reaction commm
        - for intra-reaction commm
        - for stmts
